---
title: "Popularity Index Analysis"
author: "Marc Schmieder"
date: "2/28/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
  library(data.table)
  library(checkmate)
  library(ggplot2)
  library(readxl)
  library(xgboost)
  library(extrafont)
  library(png)
  library(patchwork)
  library(devtools)
  library(elementalist)
```

```{r, include=FALSE}

# fonts
extrafont::fonts()
# extrafont::font_import()
loadfonts(device = "win")
#extrafont::fonts()
windowsFonts(Nunito = windowsFont("Nunito"))
windowsFonts(CenturyGothic = windowsFont("Century Gothic"))
windowsFonts(Ubuntu = windowsFont("Ubuntu"))


# color schemes
palette_blue <- c("#68979B",
                  "#417C81",
                  "#236267",
                  "#236267",
                  "#012F34")


palette_red <- c("#FFAAAA",
                 "#D46A6A",
                 "#AA3939",
                 "#801515",
                 "#550000")
  
palette_green <- c("#88CC88",
                   "#55AA55",
                   "#2D882D",
                   "#116611",
                   "#004400")
  
palette_brown <- c("#FFD2AA",
                   "#D49C6A",
                   "#AA6E39",
                   "#804715",
                   "#552800")


ap_scatter_plot <- function(plot_dt_long, 
                            title = "",
                            subtitle = "",
                            x = "", 
                            y = "",
                            fill = "",
                            colour = "",
                            x_col = "value",
                            y_col, 
                            fill_col = "variable", 
                            colour_col = "variable", 
                            shape_col = NULL,
                            base_size = 12,
                            round_digits_axis_x = 0,
                            round_digits_axis_y = 0) {
  assertDataTable(plot_dt_long)
  assertString(x_col)
  assertString(y_col)
  assertString(fill_col)
  if(!is.null(shape_col)) {
    assertString(shape_col)
    assertSubset(shape_col, names(plot_dt_long))
  }
  assertSubset(c(x_col, y_col), names(plot_dt_long))
  assertNumber(base_size, lower = 0)

  x_col_range <- range(na.omit(plot_dt_long[[x_col]]))
  y_col_range <- range(na.omit(plot_dt_long[[y_col]]))

  
  
  ggObj <- ggplot(plot_dt_long, aes_string(x = x_col,
                                colour = colour_col)) +
    geom_point(aes_string(y = y_col), 
               alpha = 0.8, size = 2.5) +
    # geom_linerange(data = pred_dt, aes_string(x = "data_point",
    #                                           ymin = "actuals",
    #                                           ymax = "pred_xg"),
    #                inherit.aes = FALSE, linetype = "dashed", size = 0.3,
    #                alpha = 0.5) +
    scale_x_continuous(breaks = round(seq(x_col_range[1],
                                          x_col_range[2],
                                          length.out = 10),
                                      round_digits_axis_x)) +
    scale_y_continuous(breaks = round(seq(y_col_range[1],
                                          y_col_range[2],
                                          length.out = 10),
                                      round_digits_axis_y)) +
    # scale_color_manual(values = c(actuals = palette_green[2],
    #                               pred_xg = palette_blue[2])) +
    labs(title = title,
         subtitle = subtitle,
         x = x,
         y = x,
         color = colour) +
    theme(#text = element_text(family = "Ubuntu"),
          axis.text.x = element_text(angle = 0, vjust = 0, size = base_size,
                                     margin = margin(t = 0, r = 0, 
                                                     b = 0, l = 0)),
          axis.text.y = element_text(size = base_size,
                                     margin = margin(t = 0, r = 0, 
                                                     b = 0, l = 0)),
          axis.title.x = element_text(size = base_size + 2,
                                      margin = margin(t = 10, r = 0, 
                                                      b = 0, l = 0)),
          axis.title.y = element_text(size = base_size + 2,
                                      margin = margin(t = 0, r = 10, 
                                                      b = 0, l = 0)),
          axis.ticks = element_line(colour = "gainsboro"),
          plot.title = element_text(size = base_size + 4, hjust = 0.5,
                                    margin = margin(t = 0, r = 0, 
                                                    b = 10, l = 0)),
          plot.subtitle = element_text(size = base_size , hjust = 0.5,
                                       margin = margin(t = 0, r = 0, 
                                                       b = 10, l = 0)),
          legend.background = element_rect_round(radius = unit(0.1, "snpc"), 
                                                 fill = "ghostwhite",
                                                 colour = "gainsboro", 
                                                 linetype = "solid",
                                                 size = 0.5),
          legend.title = element_text(size = base_size + 2),
          legend.text = element_text(size = base_size),
          legend.position = "top",
          plot.background = element_rect(fill = "aliceblue"),
          #panel.background = element_blank(),
          panel.background = element_rect_round(fill = "ghostwhite",
                                                colour = "gainsboro",
                                                radius = unit(0.04, "snpc"),
                                                size = 0.5, linetype = "solid"),
          #panel.background = element_blank(),
          panel.grid.major = element_line(colour = "gainsboro"),
          panel.grid.minor = element_line(colour = "gainsboro"))
  
  
  return(ggObj)
}



```


# 1. Exploration
## 1.1 Dataset attributes


```{r pressure, echo=FALSE}
dt <- fread("01_data/04_prepared_data/dt_prepared_unfiltered.csv")
```

This document summarizes a first draft from the Spotify Popularity Index (PI)
analysis.
In the current state, 331 data points from the community were provided,
whereas each data point equals one song at a given time. The data points were
anonymized by creating an artist/song ID.
Overall, 31 variables (columns) were provided or engineered through existing
data. The column names are

```{r, echo=FALSE}
names(dt)
```

As to be seen later in the analysis, the variables 
```{r}
c("StreamsLast28Days", "ListenersLast28Days",
  "StreamsLast7Days", "ListenersLast7Days")
```
are the most important ones for the later conducted model, predicting
the Popularity Index


## 1.2 Mean values for different PI

As certain PI numbers are criterias for getting into algorithmic
playlists, the following table shows the mean values in the data for
PI indices of 20-22, 30-32 and 40-42. With that information, an artist can estimate what streams/listeners/saves they need to achieve those PI's.


```{r, include=FALSE}
# means for PI ####
lower <- 20
upper <- 22
filt <- dt[between(PopularityIndex, lower, upper, incbounds = TRUE)]
aggr20 <- filt[, lapply(.SD, function(x) mean(x, na.rm = TRUE))]
res20 <- aggr20[, .(StreamsLast28Days, ListenersLast28Days,
         SavesLast28Days, StreamsLast7Days, ListenersLast7Days,
         SavesLast7Days)]
res20[, Lower := lower]
res20[, Upper := upper]

lower <- 30
upper <- 32
filt <- dt[between(PopularityIndex, lower, upper)]
aggr30 <- filt[, lapply(.SD, function(x) mean(x, na.rm = TRUE))]
res30 <- aggr30[, .(StreamsLast28Days, ListenersLast28Days,
         SavesLast28Days, StreamsLast7Days, ListenersLast7Days,
         SavesLast7Days)]
res30[, Lower := lower]
res30[, Upper := upper]



lower <- 40
upper <- 42
filt <- dt[between(PopularityIndex, lower, upper)]
aggr40 <- filt[, lapply(.SD, function(x) mean(x, na.rm = TRUE))]
res40 <- aggr40[, .(StreamsLast28Days, ListenersLast28Days,
                    SavesLast28Days, StreamsLast7Days, ListenersLast7Days,
                    SavesLast7Days)]
res40[, Lower := lower]
res40[, Upper := upper]

aggr_binded <- rbind(res20, res30, res40)

```

```{r, echo=FALSE}
aggr_binded[, .(Lower, Upper, StreamsLast28Days, ListenersLast28Days,
         SavesLast28Days, StreamsLast7Days, ListenersLast7Days,
         SavesLast7Days)]
```



## 1.3 Variables plottet vs PI


```{r, echo=FALSE}
plot_dt <- dt[, .SD, .SDcols = c("ArtistSongId",
                                      "ReleasePassed21Days",
                                       "PopularityIndex",
                                       "StreamsLast28Days",
                                       "ListenersLast28Days",
                                       "StreamsLast7Days",
                                       "ListenersLast7Days")]
plot_dt_long <- melt(plot_dt, id.vars = c("ArtistSongId",
                                          "PopularityIndex",
                                          "ReleasePassed21Days"))
plot_dt_long[, value_std := value/max(value, na.rm = TRUE),
             by = .(variable)]


ggObj <- ap_scatter_plot(plot_dt_long = plot_dt_long,
                         x_col = "value_std",
                         y_col = "PopularityIndex", 
                         colour_col = "variable",
                         shape_col = "ReleasePassed21Days",
                         x = "between 0 and 1 scaled values",
                         y = "Popularity Index",
                         title = "Scaled variable vs PI",
                         colour = "Variable",
                         base_size = 8,
                         round_digits_axis_x = 2,
                         round_digits_axis_y = 0)
ggObj

```
The grafik shows the between 0 and 1 scaled values of the Streams last 28/7 days
and listeners last 28/7 days. It can be seen that the Popularity index is 
dependent in a (not perfect) quadratic function from those variables.

# 2. Model

This is a regression problem where any model can predict continuous
numbers, but the true values can only be of integer type.
As an algorithm, the XGboost regression tree was chosen, a (still) state of the
art machine learning algorithm that builds on tree boosting.

The model was trained on 263 obversations (75 percent) and fitted on 68 
(25 percent) observations (never seen by the model). The test set of 68
data points was sampled at random, but taking into account an equal distribution
between new songs (<21 days) and older songs.

## 2.1 Model performance

```{r, echo=FALSE}
resList <- readRDS("01_data/05_model_data/model_data_MonFeb281708592022.rds")


# mse comparison
mad <- resList$metrics$MAD_xg

```
```{r, echo=FALSE}
mad
```
The mean absolute deviation (mad) of 1.52 states that the predictions of the model
deviate in mean 1.52 from the true PI. For example if the true PI for 
one song is 32, we can expect, that the model will predict a value that is
around 33.5 or 30.5. So the model is working well but not perfect. 
There can be several reasons for that the prediction of the model 
is not perfect.

1. Quality of data: false numbers entered or false PI entered
2. Time-delay of PI (updates every few days, also in Spotify for artist it
is not updated in real time)
3. Rounding of PI: Internally the formula from Spotify probably results
in a continues number (e.g. 24.32) but is then rounded for the public display.
This is sort of imperfect information that can bias the model.
4. There are factors/variables that contribute to the formula that are not 
publicly accessible/not in the community driven dataset.
5. The algorithm itself lacks potential (model tuning could increase 
performance or choosing another model like random forest or a neural net)

I personally suspect that it is a combination of 1-3 and think that 4 is 
rather unlikely.


## 2.3 Variable importance

```{r, echo=FALSE}
imp <- xgb.importance(model = resList$model)
ggObj <- ggplot(data = imp, aes(x = reorder(Feature, Gain),
                                 y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(y = "Gain in XGBoost model",
       x = "Variable",
       title = "Variable Importance in XGBoost model")
ggObj
```
The grafik shows the variable importance of the computed model. That is,
how much each variable contributes to the model. It is seen that 
streams last 28 days and listeners last 28 days are by far the most important
variables. It should be noted that other variables that are highly correlate
with StreamsLast28Days and ListenersLast28Days are important too, but the 
XGBoost model only needs one representor for those sets of variables.
It can be observed, that the number of saves do appear very late in the 
variable importance ranking.

## 2.3 Graph of predictions

Lets take a look at the predictions and true values of the test data set.

```{r, include=FALSE}
pred_dt <- resList[["pred_dt"]]
setorderv(pred_dt, "actuals", 1)
pred_dt[, data_point := 1:.N]


base_size <- 8
plotData <- melt(pred_dt, id.vars = c("data_point",
                                      "ArtistSongId"))
```


```{r, echo=FALSE}
ggObj <- ggplot(plotData, aes(x = data_point,
                              colour = variable)) +
  geom_point(aes_string(y = "value"), 
             alpha = 0.9, size = 3) +
  geom_linerange(data = pred_dt, aes_string(x = "data_point",
                            ymin = "actuals",
                            ymax = "pred_xg"),
                  inherit.aes = FALSE, linetype = "dashed", size = 0.3,
                 alpha = 0.5) +
  scale_y_continuous(breaks = seq(0, max(plotData[["value"]]), by = 5)) +
  scale_x_continuous(breaks = seq(0, max(plotData[["data_point"]]), by = 5)) +
  scale_color_manual(values = c(actuals = palette_green[2],
                                pred_xg = palette_blue[2])) +
  labs(title = "Evaluation Plot:  Comparison of predictions and true values",
       subtitle = "Model:  XGboost",
       x = "Data point in unseen data",
       y = "Popularity Index",
       color = "Type of point:") +
  theme(#text = element_text(family = "Ubuntu"),
        axis.text.x = element_text(angle = 0, vjust = 0, size = base_size,
                                   margin = margin(t = 0, r = 0, 
                                                   b = 0, l = 0)),
        axis.text.y = element_text(size = base_size,
                                   margin = margin(t = 0, r = 0, 
                                                   b = 0, l = 0)),
        axis.title.x = element_text(size = base_size + 2,
                                    margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)),
        axis.title.y = element_text(size = base_size + 2,
                                    margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0)),
        axis.ticks = element_line(colour = "gainsboro"),
        plot.title = element_text(size = base_size + 4, hjust = 0.5,
                                  margin = margin(t = 0, r = 0, 
                                                  b = 10, l = 0)),
        plot.subtitle = element_text(size = base_size , hjust = 0.5,
                                     margin = margin(t = 0, r = 0, 
                                                     b = 10, l = 0)),
        legend.background = element_rect_round(radius = unit(0.1, "snpc"), 
                                               fill = "ghostwhite",
                                               colour = "gainsboro", 
                                               linetype = "solid",
                                               size = 0.5),
        legend.title = element_text(size = base_size + 2),
        legend.text = element_text(size = base_size),
        legend.position = "top",
        plot.background = element_rect(fill = "aliceblue"),
        #panel.background = element_blank(),
        panel.background = element_rect_round(fill = "ghostwhite",
                                        colour = "gainsboro",
                                        radius = unit(0.04, "snpc"),
                                        size = 0.5, linetype = "solid"),
        #panel.background = element_blank(),
        panel.grid.major = element_line(colour = "gainsboro"),
        #panel.grid.minor = element_line(colour = "gainsboro"),
        panel.grid.minor = element_blank())
ggObj

```

The grafik shows the prediction of the XGBoost model vs the actual values.
Most observations were predicted with decent accuracy. Only the predictions
of some points show a serious deviation from the actual values. It could be 
of merit to look into those data points to see which reasons 1-5 from 2.2
do apply.








